% LaTeX source for the ordinalArithmetic document
%

\InputIfFileExists{options}{}{}
\documentclass[a4paper]{amsart}
\usepackage{disitt}
\usepackage{disitt-symbols}

\begin{document}
\sloppy

\title[A short title]{The full title}
\input{frontMatter}
\subjclass[2010]{Primary unknown; Secondary unknown} %
\keywords{Keyword one, keyword two etc.}%

\begin{abstract}
This is the abstract.
\end{abstract}
\maketitle
\tableofcontents

\section{Introduction}

Essentially, the only and ultimate goal of Mathematics, is to build
\emph{Plato's Playground}. That is the ``collection'' of DiSiTT structures which
it is possible to build. Of great use to the Mathematical Sciences, is our
collective \emph{interpretation} of a given structure in Plato's Playground.

In order to ``understand'' our mathematics at the meta-theory level, we need to
start exploring Plato's Playground by building the Ordinals, \Ordinal.

We begin at the \emph{theory} level by positing the existence of a \emph{type},
the Ordinals, \Ordinal and an \emph{an} object, \judgement{\zero}{\Ordinal}.

\begin{deAxiom}
\conclusion{}{\judgement{\Ordinal}{\Universe}}
\end{deAxiom}

\begin{deAxiom}
\conclusion{}{\judgement{\zero}{\Ordinal}}
\end{deAxiom}

Given an ordinal, $\alpha$, we can create a new ordinal in two distinctly
different ways. We can form the \emph{successor}, \successor{\alpha}:

\begin{deAxiom}
\hypothesis{\Gamma}{\judgement{\alpha}{\Ordinal}}
\conclusion{\Gamma}{\judgement{\successor{\alpha}}{\Ordinal}}
\end{deAxiom}

We can also form a new \emph{limit} ordinal, \limitOrd{\alpha}:

\begin{deAxiom}
\hypothesis{\Gamma}{\judgement{\alpha}{\Ordinal}}
\conclusion{\Gamma}{\judgement{\limitOrd{\alpha}}{\Ordinal}}
\end{deAxiom}

So far we have posited the existence of the \emph{objects}, or \zero-simplices,
in the Ordinal structure. The quintiessential property of the Ordinals, is that
they form a totally ordered structure. To express this aspect of the Ordinals,
we need to specify the ``arrows'', or \one-simplices, in the Ordinal structure.

\begin{deAxiom}
\hypothesis{\Gamma}{\judgement{\alpha}{\Ordinal}}
\conclusion{\Gamma}{
  \judgement{\arrow{\alpha}{\successor{\alpha}}}{\Ordinal}
}
\end{deAxiom}

\begin{deAxiom}
\conclusion{}{
  \judgement{\arrow{\zero}{\limitOrd{\zero}}}{\Ordinal}
}
\end{deAxiom}

\begin{deAxiom}
\hypothesis{\Gamma}{\judgement{\alpha}{\Ordinal}}
\conclusion{\Gamma}{
  \judgement{\arrow{\limitOrd{\alpha}}{\limitOrd{\successor{\alpha}}}}{\Ordinal}
}
\end{deAxiom}

\begin{deAxiom}
\hypothesis{\Gamma}{\judgement{\alpha}{\Ordinal}}
\hypothesis{\Gamma}{\judgement{\beta}{\Ordinal}}
\hypothesis{\Gamma}{\judgement{\arrow{\alpha}{\limitOrd{\beta}}}{\Ordinal}}
\conclusion{\Gamma}{
  \judgement{\arrow{\successor{\alpha}}{\limitOrd{\beta}}}{\Ordinal}
}
\end{deAxiom}

\TODO{There is an implicit assumption here that the first four axioms only deal
with objects.  That is you can not form a successor or limit-ordinal from an
arrow!! How do we make this assumption explicit? }

With the very minimal tools available to us, we would now like to show two
things.  Firstly, that every ordinal is the ``ultimate successor'' of some
identifible limit-ordinal, and secondly, that every ordinal has a one-simplex
(``arrow'') directed to the ``next'' limit-ordinal.

We define the ultimate successor ``function'' as follows:

%\begin{deDefinition}
\begin{deAxiom}
\hypothesis{\Gamma}{\judgement{\alpha}{\Ordinal}}
\conclusion{\Gamma}{
  \judgement{\arrow{\limitOrd{\alpha}}{\limitOrd{\alpha}}}{\TODO{WHAT GOES HERE?}}
}
\end{deAxiom}
%\end{deDefinition}

%\begin{deDefinition}
\begin{deAxiom}
\hypothesis{\Gamma}{\judgement{\alpha}{\Ordinal}}
\hypothesis{\Gamma}{\judgement{\beta}{\Ordinal}}
\hypothesis{\Gamma}{\judgement{\arrow{\alpha}{\successor{\alpha}}}{\Ordinal}}
\hypothesis{\Gamma}{\judgement{\arrow{\alpha}{\limitOrd{\beta}}}{\TODO{WHAT GOES HERE?}}}
\conclusion{\Gamma}{
  \judgement{\arrow{\successor{\alpha}}{\limitOrd{\beta}}}{\TODO{WHAT GOES HERE?}}
}
\end{deAxiom}
%\end{deDefinition}

We define the next limit ordinal function as follows:

%\begin{deDefinition}
\begin{deAxiom}
\hypothesis{\Gamma}{\judgement{\alpha}{\Ordinal}}
\conclusion{\Gamma}{
  \judgement{\arrow{\limitOrd{\alpha}}{\limitOrd{\successor{\alpha}}}}{\TODO{WHAT GOES HERE?!}}
}
\end{deAxiom}
%\end{deDefinition}

%\begin{deDefinition}
\begin{deAxiom}
\hypothesis{\Gamma}{\judgement{\alpha}{\Ordinal}}
\hypothesis{\Gamma}{\judgement{\beta}{\Ordinal}}
\hypothesis{\Gamma}{\judgement{\arrow{\alpha}{\successor{\alpha}}}{\Ordinal}}
\hypothesis{\Gamma}{\judgement{\arrow{\alpha}{\limitOrd{\beta}}}{\TODO{WHAT GOES HERE?!}}}
\conclusion{\Gamma}{
  \judgement{\arrow{\successor{\alpha}}{\limitOrd{\beta}}}{\TODO{WHAT GOES HERE?!}}
}
\end{deAxiom}
%\end{deDefinition}

Finally we have the axiom:
\newcommand{\nextLimitOrd}[1]{\ensuremath{\hat{\omega}(#1)}}
\begin{deAxiom}
\hypothesis{\Gamma}{\judgement{\alpha}{\Ordinal}}
\conclusion{\Gamma}{
  \judgement{\arrow{\alpha}{\nextLimitOrd{\alpha}}}{\Ordinal}
}
\end{deAxiom}

\TODO{Here be dragons! We could posit ``the concatination of two arrows is an
arrow'', but then how do we distinguish between ``atomic'' arrows and
``non-atomic'' (derived) arrows? Also to prove that every ordinal is an ultimate
successor of an identifible limit-ordinal, we need to essentially run the above
axioms in reverse (to provide a ``negative'' conclusion -- something can't
happen). We certainly need to introduce (co)induction!! Also, we really would
like $\limitOrd{\zero} = \zero$, how do we ``cleanly'' do this?}

\bibliographystyle{amsalpha}
\bibliography{ordinalArithmetic}

\end{document}

