% A ConTeXt document [master document: diSimp.tex]

\chapter[title=diSimp Copyrights]

We use the following dual Creative Commons Attribution-ShareAlike and Apache
Licenses for all documents.

\startMkIVCode
\def\placeCopyrightPage#1#2{
  \startmakeup

%\blank[10cm]
\blank[6cm]

\doifvariable{diSimp}{ISBN}{
ISBN-13: \getvariable{diSimp}{ISBN}
\blank
}

Unless explicitly stated otherwise, all content contained in this
document which is \emph{not software code} is

\blank[small]

\startcenteraligned
\bold{Copyright © \currentdate[year] #1}
\stopcenteraligned

\blank[small]

and is licensed for release under the Creative Commons
Attribution-ShareAlike 4.0 International License (CC-BY-SA 4.0 License).
You may obtain a copy of the License at

\blank[small]

\startcenteraligned
\bold{\type{http://creativecommons.org/licenses/by-sa/4.0/}}
\stopcenteraligned

\blank[small]

Unless required by applicable law or agreed to in writing, all non-code
content distributed under the above CC-BY-SA 4.0 License is distributed
on an \quote{AS IS} BASIS, WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND,
either express or implied. See the CC-BY-SA 4.0 License whose URL is
listed above for the specific language governing permissions and
limitations under the License.

\blank[medium]
\hrule
\blank[medium]

Again, unless explicitly stated otherwise, all content contained in this
document which is \emph{software code} is

\blank[small]

\startcenteraligned
\bold{Copyright © \currentdate[year] #1}
\stopcenteraligned

\blank[small]

and is licensed under the Apache License, Version 2.0 (the "License");
you may not use this code except in compliance with the License. You may
obtain a copy of the License at

\blank[small]

\startcenteraligned
\bold{\type{http://www.apache.org/licenses/LICENSE-2.0}}
\stopcenteraligned

\blank[small]

Unless required by applicable law or agreed to in writing, software
distributed under the Apache 4.0 License is distributed on an \quote{AS
IS} BASIS, WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express
or implied. See the Apache 2.0 License whose URL is listed above for the
specific language governing permissions and limitations under the
License.

\stopmakeup
}
\stopMkIVCode

We use the following Apache License for all code produced from diSimplex 
documents.

\startMkIVCode
\def\placeMkIVCopyright#1{\addApacheLicense{MkIVCode}{}{#1}}

\def\placeLuaCopyright#1{\addApacheLicense{LuaCode}{--}{#1}}

\def\placeCHeaderCopyright#1{\addApacheLicense{CHeader}{//}{#1}}
\def\placeCCodeCopyright#1{\addApacheLicense{CCode}{//}{#1}}

\def\placeJoylolCopyright#1{\addApacheLicense{JoylolCode}{//}{#1}}
\stopMkIVCode

